---
title: "HW5"
output: html_document
date: "2023-03-31"
---

```{r setup, include=FALSE}
library(tidyverse)
library(readr)
library(pscl)
```

### QUESTION 1

#### (a) Model Fitting and Interpretation
```{r}
## fit a poisson log-link model
crab = read.table("HW5-crab.txt", header = TRUE)
M1 = glm(Sa ~ W, family=poisson(link=log), data=crab)
summary(M1)

# goodness of fit
res.pearson = residuals(M1, type="pearson")
G.stat = sum(res.pearson^2)
res.dev = residuals(M1, type="deviance")
D.stat = sum(res.dev^2)
1-pchisq(G.stat, df=173-2)
1-pchisq(D.stat, df=173-2)

```
**Goodness of fit:**

The Pearson's Chi-square test statistic is `r round(G.stat, 2)`, deviance test statistic is `r round(D.stat, 2)`. The p-values of both of them are less than 0.05 under 171 df, indicating the fit is not good here.

**Interpretation of the model:**

We fitted a poisson log linear model here where:

log(E(Y)) = β0 + β1·X

with Y=Sa (number of satellites), X=W (carapace width).

Here β0 = `r round(M1$coefficients[1], 2)`, β1 =  `r round(M1$coefficients[2], 2)`. Indicating that when the carapace width is 0, the expected number of satellites is `r round(exp(M1$coefficients[1]), 2)` . When there's one unit change in carapace width, the expected change in the number of satellites is `r round(exp(M1$coefficients[2]), 2)`.


#### (b) Model Fitting with More Predictors and Interpretation
```{r}
# fit again with 2 predictors
M2 = glm(Sa ~ W + Wt, family=poisson(link=log), data=crab)
summary(M2)

# goodness of fit
res.pearson2 = residuals(M2, type="pearson")
G.stat2 = sum(res.pearson2^2)
res.dev2 = residuals(M2, type="deviance")
D.stat2 = sum(res.dev2^2)
1-pchisq(G.stat2, df=173-3)
1-pchisq(D.stat2, df=173-3)
```


```{r}
# compare 2 models
test.stat = M1$deviance - M2$deviance
df = 171-170
pv3 =  1-pchisq(test.stat, df = df)
pv3 # rej, choose larger model
```
The Pearson's Chi-square test statistic is `r round(G.stat2, 2)`, deviance test statistic is `r round(D.stat2, 2)`. The p-values of both of them are less than 0.05 under 170 df, indicating the fit is not good here. However, the p-value of the deviance test comparing the two models is less than 0.05 under 1 df, indicating that we should choose the larger model over the smaller one.


#### (c) Overdispersion Adjustment

```{r}
# plot half-normal plot
abs_res = abs(res.pearson2)
plot(qnorm((173+1:173+0.5)/(2*173+1.125)),sort(abs_res))
abline(a=0,b=1)

# estimate dispersion parameter
res.p1 = residuals(M2, type='pearson', data=crab)
G1 = sum(res.p1^2)
phi = G1/(173-3)
phi
# refit with phi
summary(M2, dispersion=phi)
# goodness of fit
pv.od = 1-pchisq(173-3, df=173-3)
```
The dispersion parameter is `r round(phi, 2)` in M2. There is a deviation from the reference line with slope 1, indicating constant over-dispersion. After adjusting for dispersion=`r round(phi, 2)`, the model parameters are the same as before, but the p-value for the Chi-square test is `r round(pv.od, 2)` here, greater than 0.05, indicating a good fit.


### QUESTION 2

#### (a) Model Fitting and Interpretation
```{r}
# fit a poisson log-link model
fish = read.table("HW5-parasite.txt", header = TRUE)
fish = 
  fish %>% 
  mutate(Area = as.factor(Area),
         Year = as.factor(Year)) 
M1.fish = glm(Intensity ~ Area + Year + Length, family=poisson(link=log), data=fish)
summary(M1.fish)
```
**Interpretation of the model:**

We fitted a poisson log linear model here where:

log(E(Y)) = β0 + β1·X1 + β2·X2 + β3·X3

with Y = Intensity, X1 = Area2, X2 = Area3, X3 = Area3, X4 = Year2000, X5 = Year2001, X5 = Length.

- Here exp(β1) = `r round(exp(M1.fish$coefficients[2]), 2)`, indicating that adjusting for year and length, the expected intensity of parasite in area2 is `r round(exp(M1.fish$coefficients[2]), 2)` times of that in area1.

- exp(β2) = `r round(exp(M1.fish$coefficients[3]), 2)`, indicating that adjusting for year and length, the expected intensity of parasite in area3 is `r round(exp(M1.fish$coefficients[3]), 2)` times of that in area1.

- exp(β3) = `r round(exp(M1.fish$coefficients[4]), 2)`, indicating that adjusting for year and length, the expected intensity of parasite in area4 is `r round(exp(M1.fish$coefficients[4]), 2)` times of that in area1.

- exp(β4) = `r round(exp(M1.fish$coefficients[5]), 2)`, indicating that adjusting for area and length, the expected intensity of parasite in year 2000 is `r round(exp(M1.fish$coefficients[5]), 2)` times of that in year 1999.

- exp(β5) = `r round(exp(M1.fish$coefficients[6]), 2)`, indicating that adjusting for area and length, the expected intensity of parasite in year 2001 is `r round(exp(M1.fish$coefficients[6]), 2)` times of that in year 1999.

- exp(β6) = `r round(exp(M1.fish$coefficients[7]), 2)`, indicating that adjusting for area and year, with one unit increase in length, the expected intensity of parasite would decrease by `r round(exp(M1.fish$coefficients[7]), 2)` times.


#### (b) Goodness of Fit

```{r}
# goodness of fit
res.pearson.fish = residuals(M1.fish, type="pearson")
G.stat.fish = sum(res.pearson.fish^2)
res.dev.fish = residuals(M1.fish, type="deviance")
D.stat.fish = sum(res.dev.fish^2)
1-pchisq(G.stat.fish, df=1184)
1-pchisq(D.stat.fish, df=1184)
```
**Goodness of fit:**

The Pearson's Chi-square test statistic is `r round(G.stat.fish, 2)`, deviance test statistic is `r round(D.stat.fish, 2)`. The p-values of both of them are less than 0.05 under 1184 df, indicating that the fit is not good here.

#### (c) Account for ZIP
```{r}
# analyze data for ZIP regression
M0 = zeroinfl(Intensity ~ Area + Year + Length, data = fish)
summary(M0)

```
**Interpretation of the parameters: **

**1. If the fish are susceptible to parasites:**

- exp(β1) = `r round(exp(M0$coefficients$count[2]), 2)`, indicating that adjusting for year and length, the expected intensity of parasite in area2 is `r round(exp(M0$coefficients$count[2]), 2)` times of that in area1.

- exp(β2) = `r round(exp(M0$coefficients$count[3]), 2)`, indicating that adjusting for year and length, the expected intensity of parasite in area3 is `r round(exp(M0$coefficients$count[3]), 2)` times of that in area1.

- exp(β3) = `r round(exp(M0$coefficients$count[4]), 2)`, indicating that adjusting for year and length, the expected intensity of parasite in area4 is `r round(exp(M0$coefficients$count[4]), 2)` times of that in area1.

- exp(β4) = `r round(exp(M0$coefficients$count[5]), 2)`, indicating that adjusting for area and length, the expected intensity of parasite in year 2000 is `r round(exp(M0$coefficients$count[5]), 2)` times of that in year 1999.

- exp(β5) = `r round(exp(M0$coefficients$count[6]), 2)`, indicating that adjusting for area and length, the expected intensity of parasite in year 2001 is `r round(exp(M0$coefficients$count[6]), 2)` times of that in year 1999.

- exp(β6) = `r round(exp(M0$coefficients$count[7]), 2)`, indicating that adjusting for area and year, with one unit increase in length, the expected intensity of parasite would decrease by `r round(exp(M0$coefficients$count[7]), 2)` times.

**2. Variables of whether fish are susceptible to parasites may depend on:**

- exp(z1) = `r round(exp(M0$coefficients$zero[2]), 2)`, indicating that adjusting for year and length, the odds ratio of having no parasite in area2 is `r round(exp(M0$coefficients$zero[2]), 2)` times of that in area1.

- exp(z2) = `r round(exp(M0$coefficients$zero[3]), 2)`, indicating that adjusting for year and length, the odds ratio of having no parasite in area3 is `r round(exp(M0$coefficients$zero[3]), 2)` times of that in area1.

- exp(z3) = `r round(exp(M0$coefficients$zero[4]), 2)`, indicating that adjusting for year and length, the odds ratio of having no parasite in area4 is `r round(exp(M0$coefficients$zero[4]), 2)` times of that in area1.

- exp(z4) = `r round(exp(M0$coefficients$zero[5]), 2)`, indicating that adjusting for area and length, the odds ratio of having no parasite in 2000 is `r round(exp(M0$coefficients$zero[5]), 2)` times of that in 1999.

- exp(z5) = `r round(exp(M0$coefficients$zero[6]), 2)`, indicating that adjusting for area and length, the odds ratio of having no parasite in 2001 is `r round(exp(M0$coefficients$zero[6]), 2)` times of that in 1999.

- exp(z6) = `r round(exp(M0$coefficients$zero[7]), 2)`, indicating that adjusting for year and area, the odds ratio of having no parasite in area2 would decrease by `r round(exp(M0$coefficients$zero[7]), 2)` times.

